%% technical_report.tex — LIST EuMINe DataBridge 2026
%% Compile: cd reports/technical && pdflatex technical_report.tex (twice)

\documentclass[10pt, a4paper]{article}
\usepackage{eumine_hackathon}

\title{%
  \textbf{Technical Report}\\[4pt]
  {\large EuMINe DataBridge Hackathon \hackathonyear}
}
\author{
  LIST\_EuMINe \\
  \small Luxembourg Institute of Science and Technology (LIST) \\
  \small euminecost@gmail.com
}
\date{May 2026}

\begin{document}
\maketitle
\thispagestyle{fancy}

\begin{abstract}
We predict formation energy per atom ($E_f$) and band gap ($E_g$) for
inorganic crystals using a weighted ensemble of fine-tuned ALIGNN graph
neural networks and MACE-MP-0 universal potentials, with isotonic
post-hoc calibration. Training on 850 Bridge Dataset structures with
5-fold cross-validation yields an honest out-of-fold score of
\textbf{35.17/40} ($E_f$ MAE: 0.053~eV/atom, $E_g$ MAE: 0.195~eV).
The predictor implements MatFed API~v1 (15/15 automated compliance tests).
\end{abstract}

\tableofcontents
\vspace{1em}

\section{Model Architecture}

\subsection{Overview}
The pipeline stacks four layers: (1)~ALIGNN models fine-tuned from
JARVIS-pretrained weights for $E_f$ and $E_g$; (2)~MACE-MP-0 (medium)
for $E_f$ via per-element reference subtraction and for $E_g$ via a
gradient-boosting head on invariant node descriptors; (3)~Optuna
weighting optimised against the hackathon score; (4)~isotonic
regression calibration. Uncertainty is reported as
$|\text{ALIGNN} - \text{MACE}|$ per property.

\subsection{Input Representation}
Crystal structures (\texttt{pymatgen.Structure}) are converted to
JARVIS atom dictionaries for ALIGNN line graphs ($k$-NN, 8~\AA\
cutoff, 12 neighbours) and to ASE atoms for MACE equivariant
message-passing.

\subsection{Architecture Details}
\begin{itemize}
  \item ALIGNN EF: 4 ALIGNN + 4 GCN layers, 256 hidden features
  \item ALIGNN BG: 6 ALIGNN + 4 GCN layers, 256 hidden features
  \item MACE-MP-0 medium; GBM BG head (300 trees, depth 4)
  \item Ensemble: two weights per property (ALIGNN vs.\ MACE)
\end{itemize}

\section{Training Procedure}

\subsection{Data Split Used}
Development: train 700 / val 150 / test 150 (labels on test withheld).
Final models: train+val combined (850) with 10\% internal holdout for
early stopping only. Five-fold stratified CV (metal / semiconductor /
wide-gap) for unbiased OOF evaluation.

\subsection{Loss Function}
Huber loss per property ($\delta = 0.5$ for $E_f$, $1.0$ for $E_g$).

\subsection{Optimiser and Hyperparameters}
\begin{table}[h!]
\centering
\caption{Training hyperparameters (ALIGNN fine-tuning).}
\label{tab:hparams}
\begin{tabular}{@{}ll@{}}
\toprule
Hyperparameter & Value \\
\midrule
Optimiser           & Adam \\
Learning rate       & $10^{-3}$ \\
Scheduler           & CosineAnnealingLR \\
Batch size          & 32 (adjusted for val divisibility) \\
Epochs (EF / BG)    & 300--350 / 400--450 \\
Early stopping      & Patience 50--70 \\
Weight decay        & $10^{-5}$ \\
Encoder freeze      & 20--30 epochs \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Hardware and Runtime}
NVIDIA RTX A4000 (16~GB); full 4B retraining $\sim$2--4~hours.

\section{Results}

\begin{table}[h!]
\centering
\caption{Performance summary (validation and OOF).}
\label{tab:results}
\begin{tabular}{@{}lcccc@{}}
\toprule
Model & MAE $E_f$ & MAE $E_g$ & Score $E_f$ & Score $E_g$ \\
      & (eV/atom) & (eV) & (/20) & (/20) \\
\midrule
Hackathon baseline & 0.238 & 0.641 & 10.0 & 10.0 \\
ALIGNN only (Step 3, val) & 0.076 & 0.198 & 17.1 & 17.0 \\
Ensemble + cal.\ (Step 4, val) & 0.052 & 0.150 & 18.2 & 17.8 \\
\textbf{OOF (5-fold CV)} & \textbf{0.053} & \textbf{0.195}
    & \textbf{18.1} & \textbf{17.1} \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textbf{Total hackathon score:} Step~4 val = 35.96/40;
OOF (honest) = \textbf{35.17/40} ($\Delta = -0.79$, expected
optimism correction).

\subsection{Results by Material Class}
Performance is stratified in cross-validation by band-gap category
(metal $E_g = 0$, semiconductor $0 < E_g < 3$~eV, wide-gap $E_g \geq 3$~eV).
Fold~4 showed elevated EF MAE ($0.082$~eV/atom) relative to
other folds ($0.061$--$0.063$), likely due to a higher
concentration of complex transition-metal oxides in that
held-out partition.

\subsection{Error Analysis}
MACE contributes $\sim$22\% weight to $E_f$ but $\sim$1\% to $E_g$
(overfit BG head on train). Isotonic calibration provides the largest
post-ensemble gain. Large $|\text{ALIGNN} - \text{MACE}|$ on individual
structures (e.g.\ sample struct\_03, $\sim$2.5~eV/atom) flags
out-of-distribution cases for federation.

\section{Federation Readiness}
\texttt{LISTEuMINePredictor} implements MatFed API~v1: 5/5 official
\texttt{test\_interface.py} tests and 10/10 project compliance tests.
Rich \texttt{describe()} metadata documents architecture, training
provenance, and uncertainty method.

\section{Reproducibility}

\subsection{Code Repository}
\texttt{eumine\_databridge} (EuMINe DataBridge 2026 repository).

\subsection{Dependencies}
Python 3.12; PyTorch; \texttt{alignn}; \texttt{mace-torch}; pymatgen;
scikit-learn; Optuna (see \texttt{requirements.txt}).

\subsection{Reproducing the Results}
\begin{lstlisting}[language=bash]
conda activate ip
cd eumine_databridge
pip install -e .
python scripts/retrain_full.py
bash scripts/run_matfed_compliance.sh
\end{lstlisting}

\section{Discussion}

\subsection{Key Findings}
Cross-source MP/JARVIS harmonisation (603 pairs, $R^2 \approx 0.96$ on
$E_f$) supports stable ALIGNN targets. OOF scoring is the unbiased
estimate for leaderboard expectation ($\sim$35.2$\pm$0.5); the 850-sample
full model used for test submission may score between OOF and Step~4 val.

\subsection{Limitations}
MACE BG head overfits small-data regimes; complex oxides remain
challenging (Fold~4). Calibration on OOF is honest but single-pass
(not nested per-fold calibrators).

\subsection{Future Work}
Nested CV for calibration; regularised MACE BG; larger pretraining
corpora; federation composition in Stage~2.

\section*{Conclusion}
We present an ALIGNN--MACE ensemble with isotonic calibration,
achieving a validation score of \textbf{35.96/40} on the fixed
150-sample validation set and an honest 5-fold cross-validated
OOF estimate of \textbf{35.17/40}
(EF MAE: $0.053$~eV/atom, BG MAE: $0.195$~eV).
Both properties substantially beat the baseline
(RF+MAGPIE: EF $0.238$, BG $0.641$~eV).
Full MatFed API v1 compliance is verified, and uncertainty
estimates are provided for all predictions.

\section*{References}
\begin{enumerate}
  \item Jain et al., \textit{APL Mater.} \textbf{1}, 011002 (2013).
  \item Choudhary and DeCost, \textit{npj Comput. Mater.} \textbf{7}, 22 (2021).
  \item Batatia et al., arXiv:2401.00096 (MACE-MP).
  \item Choudhary et al., \textit{npj Comput. Mater.} \textbf{8}, 204 (2022) (ALIGNN).
\end{enumerate}

\end{document}
